NanoMaLy

### 2015-11-12 paper/pldi16-submission/semantics.tex
proof of lem:refine-partial

\begin{itemize}
\item how do we run ill-typed programs?
\item for a lang like ocaml, dynamic semantics are independent of static
  semantics, just lambda calculus. so no problem to run ill-typed
  program
\item but what about functions? what type of arguments should we pass? consider

\begin{lstlisting}
let f x = 
  let y = 1 + x in
    1. +. y
\end{lstlisting}

does \texttt{f} take an int, float, string? int and float are both
somewhat plausible, but string or anything else is ``clearly'' bogus. so
we cannot provide \emph{completely arbitrary} inputs to
\texttt{f}. Instead, we call \texttt{f} with a \emph{hole}, written
\ehole{}, which is a placeholder for a value whose type we have not
yet determined. As we execute the program, we instantiate holes with
concrete values as demanded by the primitive operations in the
program. For example, the hole we pass to f will be instantiated to an
int when we reach the \lstinline{1 + x} term. Thus, y will be an int as
well, and the program will get stuck at \lstinline{1. +. y}. \ES{this
  reads more like overview text..}

% \item values are tagged with their types, just like ``untyped'' langs
% \item special ``hole'' value whose type is not yet known, used for function args
% \item on-the-fly unification to determine ``correct'' type for holes
\end{itemize}

Next, we formalize our search for witnesses to type-errors.
%
We present the syntax and operational semantics of \lang -- a simple
lambda calculus with integers and booleans, extended with our notion of
holes -- as well as our search algorithm.
%
We prove that our system \emph{soundly} finds witnesses, \ie if we find
a witness then there is no possible typing for the input program.
%
\subsection{Syntax}
\label{sec:syntax}
\input{syntax.tex}
%
Figure~\ref{fig:syntax} describes the syntax of \lang, a simple lambda
calculus with numbers and booleans.
%
As we are specifically interested in
programs that \emph{do} go wrong, we include an explicit \stuck state in
our syntax.
%
\paragraph{Holes}
\label{sec:holes}
The main novelty in our system is the notion of a ``hole'', written
\ehole{}, which represents an uninstantiated value.
%
Importantly, we do not even know what type the value should have.
%
Holes may also appear in types, where they may be thought of as type
variables that we will not generalize over.
%
\subsection{Semantics}
\label{sec:semantics}
\input{operational.tex}
%
Figure~\ref{fig:operational} describes the small-step contextual
reduction semantics for \lang.
%
We write \stepi{j}{e}{\su}{e'}{\su'} if there exist $e_1,\ldots,e_j$ such that
$e$ is $e_1$, $e'$ is $e_j$ and $\forall i,j, 1 \leq i < j$, we have
\step{e_i}{\su_i}{e_{i+1}}{\su_{i+1}}.
%
We write \steps{e}{\su}{e'}{\su'} if there exists some (finite) $j$ such that
$\stepi{j}{e}{\su}{e'}{\su'}$.
%
The evaluation relation is parameterized by a pair of \forcesym and
\gensym functions that handle runtime type-checking and instantiation of
holes respectively.
%
The relation must also maintain a substitution \su, mapping holes to
concrete values, so that we do not instantiate the same hole with
different values in different contexts, and so that we can report a
concrete witness to any discovered type errors.

@@

  We will omit the subscript when we wish to refer to the final partial
  type. \ES{need to define typeof..}
\end{defn}
%
We also define a compatibility relation between types.
%
\begin{defn}
\label{defn:type-compat}
  A type $s$ is \emph{compatible} with a type $t$, written \tcompat{s}{t}, if
  $\exists \su.\ t = \subst{\su}{s} \lor s = \subst{\su}{t}$.
\end{defn}
%
Given these two definitions, we show that each evaluation step
refines the partial type of $f$.
%
\begin{lem}
\label{lem:refine-partial}
  For all $k$, \tcompat{\ptype{k}{f}}{\ptype{k+1}{f}}.
\end{lem}
%
\begin{lem}
\label{lem:force-inst}
  For all $k$, if $\ptype{k}{f} \neq \ptype{k+1}{f}$, then \forcesym must
  have been called at step $k+1$.
\end{lem}
\begin{proof}
  By case analysis on the evaluation rules. 
  % 
  If $\ptype{k}{f} \neq \ptype{k+1}{f}$ then one of the holes in $f$'s
  argument must have been instantiated with a concrete value at step
  $k+1$.
  %
  An examination of the rules shows that only place this happens is 
  in the first case of \forcesym.
\end{proof}
%
Finally, we show that any value that is not compatible with the $k$th
partial type of $f$ will cause $f$ to get stuck in \emph{at most} $k$
steps.
%
\begin{lem}
\label{lem:k-stuck}
  For any \hastype{v}{t} that is not compatible with
  the input of \ptype{k}{f}, \stepi{k}{\eapp{f}{v}}{\emptysu}{\stuck}{\su}.
\end{lem}
\begin{proof}
  By induction on $k$.

  Let $s_k = \ptype{k}{f}$. Suppose \hastype{v}{\tincompat{t}{s_{k+1}}}, we
  will show that \stepi{k+1}{\eapp{f}{v}}{\emptysu}{\stuck}{\su}.
  \begin{description}
  \item[Case \tincompat{t}{s_k}:] 
    The inductive hypothesis applies.
  \item[Case \tcompat{t}{s_k} but \tincompat{t}{s_{k+1}}:] 
    By Lemma~\ref{lem:force-inst} we must have called \forcesym at step
    $k+1$. A case analysis of the applicable rules shows that \forcesym
    cannot have succeeded.
  \end{description}
\end{proof}
%
Now we can prove our soundness theorem.
%
\begin{thm}
\label{thm:soundness}
  For any function $f$, if \steps{\eapp{f}{\ehole{}}}{\emptysu}{\stuck}{\su},
  then $\forall t. \exists \hastype{v}{t}. \steps{\eapp{f}{v}}{\emptysu}{\stuck}{\su}$.
\end{thm}
\begin{proof}
  Let $\tfun{s}{\_} = \ptype{}{f}$.
  \begin{description}
  \item[Case \tcompat{s}{t}:] Our witness is already valid.
  \item[Case \tincompat{s}{t}:] We can apply Lemma~\ref{lem:k-stuck} and
    use \emph{any} \hastype{v}{t} as a witness for $f$.
  \end{description}
\end{proof}

>>>>

% \begin{itemize}
% \item how do we run ill-typed programs?
% \item for a lang like ocaml, dynamic semantics are independent of static
%   semantics, just lambda calculus. so no problem to run ill-typed
%   program
% \item but what about functions? what type of arguments should we pass? consider
%
% \begin{lstlisting}
% let f x = 
%   let y = 1 + x in
%     1. +. y
% \end{lstlisting}
%
% does \texttt{f} take an int, float, string? int and float are both
% somewhat plausible, but string or anything else is ``clearly'' bogus. so
% we cannot provide \emph{completely arbitrary} inputs to
% \texttt{f}. Instead, we call \texttt{f} with a \emph{hole}, written
% \ehole{}, which is a placeholder for a value whose type we have not
% yet determined. As we execute the program, we instantiate holes with
% concrete values as demanded by the primitive operations in the
% program. For example, the hole we pass to f will be instantiated to an
% int when we reach the \lstinline{1 + x} term. Thus, y will be an int as
% well, and the program will get stuck at \lstinline{1. +. y}. \ES{this
%   reads more like overview text..}
%
% % \item values are tagged with their types, just like ``untyped'' langs
% % \item special ``hole'' value whose type is not yet known, used for function args
% % \item on-the-fly unification to determine ``correct'' type for holes
% \end{itemize}
%
Next, we formalize our search for witnesses to type-errors.
%
We present the syntax and operational semantics of \lang -- a simple
lambda calculus with integers and booleans, extended with our notion of
holes -- as well as our search algorithm.
%
We prove that our system \emph{soundly} finds witnesses, \ie if we find
a witness then there is no possible typing for the input program.
%
\subsection{Syntax}
\label{sec:syntax}
\input{syntax.tex}
%
Figure~\ref{fig:syntax} describes the syntax of \lang, a simple lambda
calculus with numbers and booleans.
%
As we are specifically interested in
programs that \emph{do} go wrong, we include an explicit \stuck state in
our syntax.
%
\paragraph{Holes}
\label{sec:holes}
The main novelty in our system is the notion of a ``hole'', written
\ehole{}, which represents an uninstantiated value.
%
Importantly, we do not even know what type the value should have.
%
Holes may also appear in types, where they may be thought of as type
variables that we will not generalize over.
%
\subsection{Semantics}
\label{sec:semantics}
\input{operational.tex}
%
Figure~\ref{fig:operational} describes the small-step contextual
reduction semantics for \lang.
%
We write \stepi{j}{e}{\su}{e'}{\su'} if there exist $e_1,\ldots,e_j$ such that
$e$ is $e_1$, $e'$ is $e_j$ and $\forall i,j, 1 \leq i < j$, we have
\step{e_i}{\su_i}{e_{i+1}}{\su_{i+1}}.
%
We write \steps{e}{\su}{e'}{\su'} if there exists some (finite) $j$ such that
$\stepi{j}{e}{\su}{e'}{\su'}$.
%
The evaluation relation is parameterized by a pair of \forcesym and
\gensym functions that handle runtime type-checking and instantiation of
holes respectively.
%
The relation must also maintain a substitution \su, mapping holes to
concrete values, so that we do not instantiate the same hole with
different values in different contexts, and so that we can report a
concrete witness to any discovered type errors.

@@

  We will omit the subscript when we wish to refer to the final partial
  type. \ES{need to define typeof..}
\end{defn}
%
We also define a compatibility relation between types.
%
\begin{defn}
\label{defn:type-compat}
  A type $s$ is \emph{compatible} with a type $t$, written \tcompat{s}{t}, if
  $\exists \su.\ t = \subst{\su}{s} \lor s = \subst{\su}{t}$.
\end{defn}
%
Given these two definitions, we show that each evaluation step
refines the partial type of $f$.
%
\begin{lem}
\label{lem:refine-partial}
  For all $k$, \tcompat{\ptype{k}{f}}{\ptype{k+1}{f}}.
\end{lem}
\begin{proof}
  By case analysis on the evaluation rules. 
  %
  Note that all rules are type-preserving with the exception of when
  \forcesym is called on a hole, which case we instantiate the hole with
  a concrete value.
  %
  But \hastype{\ehole{}}{\thole{}}, which is compatible with any type.
\end{proof}
%
Furthermore, only a call to \forcesym can change the partial type of $f$.
%
\begin{lem}
\label{lem:force-inst}
  For all $k$, if $\ptype{k}{f} \neq \ptype{k+1}{f}$, then \forcesym must
  have been called at step $k+1$.
\end{lem}
\begin{proof}
  By case analysis on the evaluation rules. 
  % 
  If $\ptype{k}{f} \neq \ptype{k+1}{f}$ then one of the holes in $f$'s
  argument must have been instantiated with a concrete value at step
  $k+1$.
  %
  An examination of the rules shows that only place this happens is 
  in the first case of \forcesym.
\end{proof}
%
Finally, we show that any value that is not compatible with the $k$th
partial type of $f$ will cause $f$ to get stuck in \emph{at most} $k$
steps.
%
\begin{lem}
\label{lem:k-stuck}
  For any \hastype{v}{t} that is not compatible with
  the input of \ptype{k}{f}, \stepi{k}{\eapp{f}{v}}{\emptysu}{\stuck}{\su}.
\end{lem}
\begin{proof}
  By induction on $k$.

  Let $s_k = \ptype{k}{f}$. Suppose \hastype{v}{\tincompat{t}{s_{k+1}}}, we
  will show that \stepi{k+1}{\eapp{f}{v}}{\emptysu}{\stuck}{\su}.
  \begin{description}
  \item[Case \tincompat{t}{s_k}:] 
    The inductive hypothesis applies.
  \item[Case \tcompat{t}{s_k} but \tincompat{t}{s_{k+1}}:] 
    By Lemma~\ref{lem:force-inst} we must have called \forcesym at step
    $k+1$. 
    %
    A case analysis of the applicable rules shows that \forcesym cannot
    have succeeded.
  \end{description}
\end{proof}
%
Now we can prove our soundness theorem.
%
\begin{thm}
\label{thm:soundness}
  For any function $f$, if \steps{\eapp{f}{\ehole{}}}{\emptysu}{\stuck}{\su},
  then $\forall t. \exists \hastype{v}{t}. \steps{\eapp{f}{v}}{\emptysu}{\stuck}{\su}$.
\end{thm}
\begin{proof}
  Let $\tfun{s}{\_} = \ptype{}{f}$.
  \begin{description}
  \item[Case \tcompat{s}{t}:] Our witness is already valid.
  \item[Case \tincompat{s}{t}:] We can apply Lemma~\ref{lem:k-stuck} and
    use \emph{any} \hastype{v}{t} as a witness for $f$.
  \end{description}
\end{proof}
